Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-19 paper-amia-summit-2018/amia_summit_2018.tex
update introduction
update HMM model

Our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level. In this study, we utilized under and oversampling methods at data processing level for balancing the adolescent obesity data set. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.

@@

\textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. Mathematically, HMM can be defined as the following: $\lambda = (A, B, \pi)$.
\begin{itemize}
\item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
\item B is an $N\times M$ matrix $B = \{b_j(k)\}$ with observation symbol probability distribution for each state 
\item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
\end{itemize}
Hence, N is a number of hidden states in the model and M is a number of distinct observations symbols per state, i.e. the discrete vocabulary size. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.

>>>>

Our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level. In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity data set. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.

@@

\textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. Mathematically, HMM can be defined as the following: $\lambda = (A, B, \pi)$.
\begin{itemize}
\item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
\item B is an $N\times M$ matrix $B = \{b_j(k)\}$ with observation symbol probability distribution for each state 
\item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
\end{itemize}
Hence, $N$ is a number of hidden states in the model and $M$ is a number of distinct observations symbols per state, i.e. the discrete vocabulary size. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.
